Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. edited vulnerability curve results

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

<!-- # For html -->
<!-- \begin{align*} -->
<!-- \overline{O} = \frac{1}{6}\sum RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\ -->
<!-- + RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4~m}, W_{liq,~`r soil.depths[6]`~m}) + RMSE^{*}(\theta_{1~m}, W_{liq,~`r soil.depths[8]`~m}) -->
<!-- (\#eq:obj-func) -->
<!-- \end{align*} -->

@@

For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_\textrm{leaf}$ vs. $\Psi$~leaf~, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; $p$ `r  afit.p`; Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. $R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves)). Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations using Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). 
Fig. \@ref(fig:ab-fit) here.

@@

The best model (with a mulicative effect of VPD and not LAI; Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs. `r length(erd.sp)` species for the best model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, best model (with VPD and not LAI, Eq. \@ref(eq:M-chosen)).

@@

\
\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

@@

The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). WSG alone explained `r round(b.wsg.r2*100,0)`% of variation in $B$ ($p$ `r  b.wsg.p`). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

>>>>

---
title: "Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

@@

Parameters of leaf vulnerability curves were predictable from WSG and LMA. WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; $p$ `r  afit.p`; Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variance in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. $R$^2^ = `r bfit.r2`; $p$ `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves)). This predictive power allowed us to estimate leaf vulnerability curve parameters for `r sp.n.model` (out of `r length(erd.sp)` ERD species) that lacked direct observations; Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). These vulenrability curves were used in the models for effective rooting depth.

Fig. \@ref(fig:ab-fit) here.

@@

The chosen model (with a mulicative effect of VPD and not LAI; Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs. `r length(erd.sp)` species for the chosen model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (with VPD and not LAI, Eq. \@ref(eq:M-chosen)).

@@

\
\   We benefited from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

@@

The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
